\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-12-06 smurf_paper.tex
minor edits from Douglas, including a change in date format: YYYY-MM-DD

The best existing map-making strategies for recovering information on
all angular scales are maximum likelihood techniques, in which the
time-series data are expressed as a sampling of the ``true'' map of
the sky plus noise, and then the equation is inverted to estimate the
map as some weighted linear combination of the data that minimises the
variance.  The first good description of this method appears in
\citet{janssen1992} for the production of maps from the COsmic
Background Explorer (COBE -- the description is relevant despite the
fact that it used a differential radiometer instead of
bolometers). Other descriptions in the experimental CMB literature
include \citet{tegmark1997} and \citet{stompor2002}, and an
application to data from SCUBA is described in \citet{borys2004}.  The
downside to such methods is that they can be both computationally
expensive and have large memory requirements. While for some
experimental designs fast iterative methods for the inversion do exist
without requiring excessive amounts of memory, such as that of
\citet{wright1996} \citep[which was later implemented for SCUBA
by][]{johnstone2000}, for the more general map-making problem,
involving many detectors, matters are significantly complicated both
by the need to measure the cross spectra for all unique pairs of
bolometers, as well as performing the inversion itself. A
maximum-likelihood method was successfully used to make maps from ACT
data using approximately the same number of detectors as a single
\scuba\ subarray at a single wavelength. However, the calculation is
extremely resource intensive and requires a large cluster in order to
run; a single map can take 10\,CPU\,years \citep{fowler2010}. Perhaps
the most promising maximum-likelihood method that may one day be
applied to \scuba\ data is ``SANEPIC'', which produced maps from
Balloon-borne Large-Aperture Submillimeter Telescope data, involving
hundreds of bolometers, while correctly incorporating inter-bolometer
noise correlations \citep[][]{patanchon2008}.

@@

This paper is organised as follows. We first describe the properties
of \scuba\ data, including a principal component analysis to reveal
correlated noise features, in Section~\ref{sec:data}. Next, the
details of the SMURF algorithm (pre-processing steps and the iterative
solution) are given in Section~\ref{sec:algorithm}. The paper is
concluded in Section~\ref{sec:examples} with three detailed test cases
that span the majority of observation types likely to be undertaken
with \scuba, with an emphasis on the mitigation of divergence
problems, and characterising the output maps: (i) Uranus, a bright,
compact source (Section~\ref{sec:point}); (ii) the Lockman Hole, a
blind survey of faint point-like sources (Section~\ref{sec:cosmo});
and (iii) the star-forming region M17, including bright, extended
emission (Section~\ref{sec:extended}). All of the data analysed in
this paper are publicly available through the CADC \scuba\ raw-data
queries page for the dates and observation numbers given in the
text\footnote{\url{http://www.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/jcmt/search/scuba2}}.
All of the analysis was performed using the Starlink \textsc{kapuahi}
release from 2012.

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labeled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers (with a response
time constant of $\tau \sim 1$\,ms) coupled to superconducting
transition-edge sensors (TESs) for thermometry. Temperature variations
in the TESs produce changing currents, and therefore varying magnetic
fields, which are detected and amplified using chains of
superconducting quantum interference devices (SQUIDs), before the
larger output currents are digitised. Each detector has its own SQUID
for the first-stage of the amplification, but the remaining stages
occur within a common chain of SQUIDs for each of the 32 columns. All
40 rows are read out in sequence at a row-visit frequency of about
12\,kHz. Such a high sample rate is unnecessary to produce maps, so
the data are re-sampled to approximately 200\,Hz before writing to
disk. This rate provides a sample every 3.0\,arcsec, or approximately
one third of the 450\,\micron\ diffraction-limited full-width at
half-maximum (FWHM) -- a typical rule-of-thumb for adequately sampling
a Gaussian point spread function (PSF) -- at the maximum scanning
speed of 600\,arcsec\,s$^{-1}$. There is an additional 41st row of
SQUID readouts that are not connected to TESs. These ``dark SQUIDs''
track non-thermal noise sources that are common to each column's
amplifier chain. The relationship between the output digitised
current, $I$, and the input power, $P$, or $dI/dP$ is established
using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the 183\,GHz (1.6\,mm)
JCMT Water Vapour Monitor to track line-of-sight opacity variations,
see Section~3 in][]{dempsey2012}, and the application of a flux
conversion factor (FCF) which is established from regular measurements
of calibrators such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the JCMT water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  (a) Data taken during the S2SRO period,
  observation 29 on 20100313. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. (b) Data taken with the fully-commissioned instrument,
  observation 38 on 20111112. Unlike the S2SRO data, there is no
  strong signal produced by variations in the fridge temperature. The
  450\,\micron\ data have minimal correlated low-frequency noise (as
  evidenced by the lack of a difference between the raw and
  common-mode subtracted signals), although the 850\,\micron\ data
  show a common signal correlated with the opacity and telescope
  motion (note that this scan has a significantly larger amplitude
  than the S2SRO data set), mostly likely caused by a combination of
  changing airmass (anti-correlated with opacity), and magnetic field
  pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

\caption{Bolometer power spectral densities (PSDs) for the same two
  data sets (before and after upgrades) used in
  Fig.~\ref{fig:bolos_mix}. The PSDs have been boxcar smoothed with a
  width of 0.1\,Hz to reduce the noise slightly and clarify some
  features. Four of the most sensitive bolometers have been selected
  at each wavelength, and the flat-fielded and step-corrected (but
  otherwise raw) PSDs are shown as coloured dotted lines (the blue
  signals are for the same time series as those shown in
  Fig.~\ref{fig:bolos_mix}).  The solid black lines are the PSDs of
  the common-mode signals at each wavelength, and the solid coloured
  lines show the PSDs of the bolometers once the common-mode is
  removed.  Finally, the dashed black lines show the spectral shape
  produced by a point source given the scan speeds for the two
  observations (120\,arcsec\,s$^{-1}$ in (a), and
  190\,arcsec\,s$^{-1}$ in (b); for reference, the top horizontal axes
  shows the conversion from frequency to angular scale. Horizontal
  dotted lines at $10^{-7}$ and $10^{-8}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, are provided as a visual reference for
  the white-noise levels. In addition to $1/f$ and white noise
  components, and line features, all of the PSDs exhibit the gradual
  roll-off of the anti-aliasing filter just below the Nyquist
  frequency.  (a) For the S2SRO data, particularly at 450\,\micron,
  there are broad line features in the PSDs at both wavelengths above
  $\sim$35\,Hz.  At lower frequencies, the bolometer signals exhibit
  clear $1/f$ knees at approximately 2 and 3\,Hz at 450 and
  850\,\micron. Common-mode subtraction removes most of the correlated
  fridge oscillation signal, lowering the $1/f$ knees to approximately
  0.6 and 1.0\,Hz at 450 and 850\,\micron, respectively. (b) Data from
  the fully-commissioned instrument tend to have lower $1/f$ knees,
  and fewer line features. For this example, the improvement is most
  striking at 450\,micron\ where the PSD is approximately two
  orders-of-magnitude lower in the fully-commissioned data at
  0.1\,Hz. The white noise performance, however, is similar for the
  two subarrays (s4a and s8b) that were used both before and after the
  upgrades.}
\label{fig:pspec}
\end{figure*}

@@

In the S2SRO data (Fig.~\ref{fig:bolos_mix}a), observation 29 on
20100313, both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. Note that the total power in the fluctuations at
450\,\micron\ are comparable to those at 850\,\micron\ as one might
expect if there is a comparable varying thermal load from the fridge
at each wavelength that dominates. We also note that there is no
obvious strong correlation between the low-frequency signal structure,
at either wavelength, with the telescope motion. However, there is a
suggestion that the shorter-timescale behaviour of the opacity is
anti-correlated with the elevation, as expected.

@@

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. This ``scan-synchronous'' noise is
attributed to a combination of magnetic field pickup, as described in
Section~\ref{sec:magpickup}, and sky brightness variations due to
changes in elevation (not to be confused with underlying changes in
the atmosphere; a homogeneous atmosphere will appear brighter with
increasing airmass). Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

The dotted coloured lines in Fig.~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded and step-corrected (Section~\ref{sec:steps})
data. At each wavelength, and in both data sets, there are clear $1/f$
knees at frequencies ranging from roughly 1 to 2\,Hz, followed by a
predominantly white spectrum punctuated by line features at higher
frequencies, and finally a roll-off caused by the anti-aliasing filter
above $\gsim 70$\,Hz. The correlation between the low-frequency
components of the different bolometer signals is large. The solid
black lines in Fig.~\ref{fig:pspec} indicate the PSDs of the
common-modes $\mathbf{c}(t)$ at each wavelength, which reproduce much
of the low-frequency structure, as well as some of the
higher-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by uncorrelated white
noise in that part of the spectrum. The common-mode signals are fit to
each bolometer time series and removed as in Fig.~\ref{fig:bolos_mix},
and the resulting PSDs are shown as solid coloured lines.  For
reference, the top horizontal axes have been converted to angular
scale assuming the scan speeds of each observation
(120\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}a, and
190\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}b). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalised), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features at these scan speeds (this may not be the
case at higher scan speeds).

@@

Raw data from the fully-commissioned instrument
(Fig.~\ref{fig:pspec}b) generally have less significant $1/f$ noise as
compared with the S2SRO data, and it is considerably less correlated
amongst detectors (larger spread in the dotted coloured lines,
particularly noticeable in these 450\,\micron\ data), leading to a
less drastic improvement upon common-mode removal. However, since the
data are less dominated by low-frequency noise to begin with, the
signals generally require less aggressive high-pass filtering to
produce maps, and therefore retain larger-scale structures than with
the S2SRO data. Furthermore, there are generally fewer line features
in the PSDs of bolometers in the fully-commissioned
instrument. Finally, these two data sets illustrate that the white
noise performance (NEFD) is in fact similar before and after the
upgrades for these two subarrays (s4a at 450\,micron, and s8b at
850\,\micron). The main improvements are a reduction in correlated and
line noise features mentioned above, the larger number of working
bolometers and field-of-view (approximately a factor of 4).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 20100228 at 450\,\micron\ (s4a subarray).  The top
  panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smooth (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smooth. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 20100228) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a column, giving it a distinct
signature from other common signals that always appear with the same
sign.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{skynoise}
\caption{Illustration of the dominance of atmospheric variations on
  long-timescales using 35\,min. of data taken in highly-variable
  weather conditions (line-of-sight 450\,\micron\ opacity ranging from
  3.6 to 7.7) from the fully-commissioned instrument, observation 18
  on 20110324. Red and blue lines show the 850 and 450\,\micron\
  common-mode signals, corresponding to the left- and right- vertical
  axes, respectively. The black line shows the line-of-sight opacity
  from the WVM as a proxy for atmospheric emission (arbitrary
  scale). Bolometer baseline drifts on timescales $\gsim$100\,s are
  clearly correlated with the atmosphere in this case.}
\label{fig:skynoise}
\end{figure}

The data shown in Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec} were
taken during fairly stable weather conditions (although with different
450\,\micron\ line-of-sight opacities $\sim$0.9 and 6.5 in the S2SRO
and post-upgrade data sets, respectively), and are dominated at
low-frequencies ($\lsim$2\,Hz) by scan-synchronous noise, rather than
underlying variations in the atmosphere, or ``sky noise''. However,
only $\sim$100\,s of data were analysed. In Fig.~\ref{fig:skynoise},
an order-of-magnitude longer (fully-commissioned) data set is plotted
from observation 18 on 20110324, during highly variable weather
conditions (450\,\micron\ line-of-sight opacities ranging from 3.6 to
7.7, scaled from the WVM data which are shown as a black line). Unlike
the earlier plots, it is clear that the bolometer baseline drifts on
timescales $\gsim$100\,s are tightly correlated with the atmospheric
opacity, which serves as a proxy for atmospheric emission. The weaker
correlation between the WVM and the 450\,micron\ bolometers is
probably due to the atmosphere being more transmissive at 850\micron,
therefore the emission detected at 450\,\micron\ comes from water
vapour that is closer to the telescope.

While the relative importance of sky noise at low-frequencies is
highly time- and scan pattern-dependent, in general it does not have a
major impact on map-making. Upon common-mode removal (which is an
integral part of our map-making strategy) uncorrelated noise remains
(e.g., solid lines in Fig~\ref{fig:pspec}). As we will see in
Section~\ref{sec:pca} and \ref{sec:experimental}, this remaining noise
does not have a smoothly-varying correlation pattern across the focal
plane one might expect if it were due to resolved ``clouds'' of
atmospheric emission. A high-pass filter with an edge frequency in the
range $\sim$0.1--1.0\,Hz is usually employed remove this residual
noise (Section~\ref{sec:components}), which will also remove any sky
noise that is not already subtracted by the common-mode.

@@

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of
magnetic-field pickup and brightness variations in the atmosphere)?
First, the smoothly-varying gradients in the eigenvalue maps
(primarily at 850\,\micron) suggests that this noise has a different
response across the focal plane; in this case the PCA has identified
two orthogonal shapes that, when mixed in different quantities, can
reproduce most of the signal in each bolometer. It is also likely that
the atmospheric contribution is stronger at 450\,\micron\ (i.e.,
predominantly Component~3).

>>>>

The best existing map-making strategies for recovering information on
all angular scales are maximum likelihood techniques, in which the
time-series data are expressed as a sampling of the ``true'' map of
the sky plus noise, and then the equation is inverted to estimate the
map as some weighted linear combination of the data that minimizes the
variance.  The first good description of this method appears in
\citet{janssen1992} for the production of maps from the COsmic
Background Explorer (COBE -- the description is relevant despite the
fact that it used a differential radiometer instead of
bolometers). Other descriptions in the experimental CMB literature
include \citet{tegmark1997} and \citet{stompor2002}, and an
application to data from SCUBA is described in \citet{borys2004}.  The
downside to such methods is that they can be both computationally
expensive and have large memory requirements. While for some
experimental designs fast iterative methods for the inversion do exist
without requiring excessive amounts of memory, such as that of
\citet{wright1996} \citep[which was later implemented for SCUBA
by][]{johnstone2000}, for the more general map-making problem,
involving many detectors, matters are significantly complicated both
by the need to measure the cross spectra for all unique pairs of
bolometers, as well as performing the inversion itself. A
maximum-likelihood method was successfully used to make maps from ACT
data using approximately the same number of detectors as a single
\scuba\ subarray at a single wavelength. However, the calculation is
extremely resource intensive and requires a large cluster in order to
run; a single map can take 10\,CPU\,years \citep{fowler2010}. Perhaps
the most promising maximum-likelihood method that may one day be
applied to \scuba\ data is ``SANEPIC'', which produced maps from
Balloon-borne Large-Aperture Submillimeter Telescope data, involving
hundreds of bolometers, while correctly incorporating inter-bolometer
noise correlations \citep[][]{patanchon2008}.

@@

This paper is organized as follows. We first describe the properties
of \scuba\ data, including a principal component analysis to reveal
correlated noise features, in Section~\ref{sec:data}. Next, the
details of the SMURF algorithm (pre-processing steps and the iterative
solution) are given in Section~\ref{sec:algorithm}. The paper is
concluded in Section~\ref{sec:examples} with three detailed test cases
that span the majority of observation types likely to be undertaken
with \scuba, with an emphasis on the mitigation of divergence
problems, and characterising the output maps: (i) Uranus, a bright,
compact source (Section~\ref{sec:point}); (ii) the Lockman Hole, a
blind survey of faint point-like sources (Section~\ref{sec:cosmo});
and (iii) the star-forming region M17, including bright, extended
emission (Section~\ref{sec:extended}). All of the data analysed in
this paper are publicly available through the CADC \scuba\ raw-data
queries page for the dates and observation numbers given in the
text\footnote{\url{http://www.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/jcmt/search/scuba2}}.
All of the analysis was performed using the Starlink \textsc{kapuahi}
release from 2012.

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labelled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers (with a response
time constant of $\tau \sim 1$\,ms) coupled to superconducting
transition-edge sensors (TESs) for thermometry. Temperature variations
in the TESs produce changing currents, and therefore varying magnetic
fields, which are detected and amplified using chains of
superconducting quantum interference devices (SQUIDs), before the
larger output currents are digitised. Each detector has its own SQUID
for the first-stage of the amplification, but the remaining stages
occur within a common chain of SQUIDs for each of the 32 columns. All
40 rows are read out in sequence at a row-visit frequency of about
12\,kHz. Such a high sample rate is unnecessary to produce maps, so
the data are re-sampled to approximately 200\,Hz before writing to
disk. This rate provides a sample every 3.0\,arcsec, or approximately
one third of the 450\,\micron\ diffraction-limited full-width at
half-maximum (FWHM) -- a typical rule-of-thumb for adequately sampling
a Gaussian point spread function (PSF) -- at the maximum scanning
speed of 600\,arcsec\,s$^{-1}$. There is an additional 41st row of
SQUID readouts that are not connected to TESs. These ``dark SQUIDs''
track non-thermal noise sources that are common to each column's
amplifier chain. The relationship between the output digitised
current, $I$, and the input power, $P$, or $dI/dP$ is established
using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the 183\,GHz (1.6\,mm)
JCMT Water Vapour Monitor to track line-of-sight opacity variations,
see Section~3 in][]{dempsey2012}, and the application of a flux
conversion factor (FCF) which is established from regular measurements
of calibrators such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

\caption{A comparison between single bolometer time-series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature,
  the line-of-sight 450\,\micron\ opacity derived from the JCMT water
  vapour monitor, and azimuth/elevation pointing offsets, before and
  after upgrading the instrument. The grey signals over-plotted in the
  top two panels show the residual time-series after removing the
  common-mode signals.  (a) Data taken during the S2SRO period,
  observation 29 on 2010-03-13. There is a strong correlation between
  the bolometers and the roughly $\sim$25\,s oscillation in the
  fridge, but only a minor correlation with the opacity and telescope
  motion. The nearly flat common-mode subtracted signals show: (i)
  that most of the low-frequency signal is common to all of the
  bolometers; and (ii) the non-correlated, and predominantly white
  noise at 450\,\micron\ is significantly larger than at
  850\,\micron. (b) Data taken with the fully-commissioned instrument,
  observation 38 on 2011-11-12. Unlike the S2SRO data, there is no
  strong signal produced by variations in the fridge temperature. The
  450\,\micron\ data have minimal correlated low-frequency noise (as
  evidenced by the lack of a difference between the raw and
  common-mode subtracted signals), although the 850\,\micron\ data
  show a common signal correlated with the opacity and telescope
  motion (note that this scan has a significantly larger amplitude
  than the S2SRO data set), mostly likely caused by a combination of
  changing airmass (anti-correlated with opacity), and magnetic field
  pickup (see Section~\ref{sec:magpickup} and
  Fig.~\ref{fig:magpickup}).}
\label{fig:bolos_mix}
\end{figure*}

@@

\caption{Bolometer power spectral densities (PSDs) for the same two
  data sets (before and after upgrades) used in
  Fig.~\ref{fig:bolos_mix}. The PSDs have been boxcar smoothed with a
  width of 0.1\,Hz to reduce the noise slightly and clarify some
  features. Four of the most sensitive bolometers have been selected
  at each wavelength, and the flat-fielded and step-corrected (but
  otherwise raw) PSDs are shown as coloured dotted lines (the blue
  signals are for the same time series as those shown in
  Fig.~\ref{fig:bolos_mix}).  The solid black lines are the PSDs of
  the common-mode signals at each wavelength, and the solid coloured
  lines show the PSDs of the bolometers once the common-mode is
  removed.  Finally, the dashed black lines show the spectral shape
  produced by a point source given the scan speeds for the two
  observations (120\,arcsec\,s$^{-1}$ in (a), and
  190\,arcsec\,s$^{-1}$ in (b); for reference, the top horizontal axes
  shows the conversion from frequency to angular scale. Horizontal
  dotted lines at $10^{-7}$ and $10^{-8}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, are provided as a visual reference for
  the white-noise levels. In addition to $1/f$ and white noise
  components, and line features, all of the PSDs exhibit the gradual
  roll-off of the anti-aliasing filter just below the Nyquist
  frequency.  (a) For the S2SRO data, particularly at 450\,\micron,
  there are broad line features in the PSDs at both wavelengths above
  $\sim$35\,Hz.  At lower frequencies, the bolometer signals exhibit
  clear $1/f$ knees at approximately 2 and 3\,Hz at 450 and
  850\,\micron. Common-mode subtraction removes most of the correlated
  fridge oscillation signal, lowering the $1/f$ knees to approximately
  0.6 and 1.0\,Hz at 450 and 850\,\micron, respectively. (b) Data from
  the fully-commissioned instrument tend to have lower $1/f$ knees,
  and fewer line features. For this example, the improvement is most
  striking at 450\,\micron\ where the PSD is approximately two
  orders-of-magnitude lower in the fully-commissioned data at
  0.1\,Hz. The white noise performance, however, is similar for the
  two subarrays (s4a and s8b) that were used both before and after the
  upgrades.}
\label{fig:pspec}
\end{figure*}

@@

In the S2SRO data (Fig.~\ref{fig:bolos_mix}a), observation 29 on
2010-03-13, both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. Note that the total power in the fluctuations at
450\,\micron\ are comparable to those at 850\,\micron\ as one might
expect if there is a comparable varying thermal load from the fridge
at each wavelength that dominates. We also note that there is no
obvious strong correlation between the low-frequency signal structure,
at either wavelength, with the telescope motion. However, there is a
suggestion that the shorter-timescale behaviour of the opacity is
anti-correlated with the elevation, as expected.

@@

Data from the fully commissioned instrument, observation 38 on
2011-11-12, are shown in Fig.~\ref{fig:bolos_mix}b. Having solved the
fridge oscillation problem, these data no longer exhibit a correlation
with the mixing chamber (however, note that variations are still seen
in the mixing chamber signal; such variations do not necessarily
reflect changes in the focal plane temperature). There is minimal
correlated low-frequency noise in these 450\,\micron\ data, resulting
in little difference between the raw and common-mode subtracted
data. The 850\,\micron\ channel, however, exhibits a more significant
signal that is obviously correlated with the telescope motion. It is
also correlated amongst many of the detectors, and common-mode removal
corrects it to a large extent. Part of the reason that this is seen
here, and not in the S2SRO data set shown, is that the amplitude of
the scan pattern is larger. This ``scan-synchronous'' noise is
attributed to a combination of magnetic field pickup, as described in
Section~\ref{sec:magpickup}, and sky brightness variations due to
changes in elevation (not to be confused with underlying changes in
the atmosphere; a homogeneous atmosphere will appear brighter with
increasing airmass). Also similar to the S2SRO data, an
anti-correlation between the opacity and elevation is apparent.

@@

The dotted coloured lines in Fig.~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded and step-corrected (Section~\ref{sec:steps})
data. At each wavelength, and in both data sets, there are clear $1/f$
knees at frequencies ranging from roughly 1 to 2\,Hz, followed by a
predominantly white spectrum punctuated by line features at higher
frequencies, and finally a roll-off caused by the anti-aliasing filter
above $\gsim 70$\,Hz. The correlation between the low-frequency
components of the different bolometer signals is large. The solid
black lines in Fig.~\ref{fig:pspec} indicate the PSDs of the
common-modes $\mathbf{c}(t)$ at each wavelength, which reproduce much
of the low-frequency structure, as well as some of the
higher-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high frequency,
as expected if the bolometers are dominated by uncorrelated white
noise in that part of the spectrum. The common-mode signals are fit to
each bolometer time series and removed as in Fig.~\ref{fig:bolos_mix},
and the resulting PSDs are shown as solid coloured lines.  For
reference, the top horizontal axes have been converted to angular
scale assuming the scan speeds of each observation
(120\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}a, and
190\,arcsec\,s$^{-1}$ in Fig.~\ref{fig:pspec}b). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalised), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features at these scan speeds (this may not be the
case at higher scan speeds).

@@

Raw data from the fully-commissioned instrument
(Fig.~\ref{fig:pspec}b) generally have less significant $1/f$ noise as
compared with the S2SRO data, and it is considerably less correlated
amongst detectors (larger spread in the dotted coloured lines,
particularly noticeable in these 450\,\micron\ data), leading to a
less drastic improvement upon common-mode removal. However, since the
data are less dominated by low-frequency noise to begin with, the
signals generally require less aggressive high-pass filtering to
produce maps, and therefore retain larger-scale structures than with
the S2SRO data. Furthermore, there are generally fewer line features
in the PSDs of bolometers in the fully-commissioned
instrument. Finally, these two data sets illustrate that the white
noise performance (NEFD) is in fact similar before and after the
upgrades for these two subarrays (s4a at 450\,\micron, and s8b at
850\,\micron). The main improvements are a reduction in correlated and
line noise features mentioned above, the larger number of working
bolometers and field-of-view (approximately a factor of 4).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 2010-02-28 at 450\,\micron\ (s4a subarray).  The top
  panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smooth (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smooth. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

Example data from the 450\,\micron\ subarray s4a where pickup appears
to be significant (observation 16 on 2010-02-28) are shown in
Fig.~\ref{fig:magpickup}. The time-series for two bolometers in the
same column (not flatfielded) show that there is a strong signal with
a similar shape, but opposite signs. This behaviour is seen across the
entire array. The dark squid signal for the same columns exhibits a
similar shape and amplitude. Since the dark squid has no thermal
absorber or TES attached to it, this observed signal is not likely to
be optical or thermal in nature (although there can be some cross-talk
with the bolometers). Due to the fact that the sign of the gain in
each stage of SQUID amplification is random (although the combined
gain is constrained to be negative), and since magnetic field pickup
is only seen at the input to the second stage, the pickup can appear
with random signs for bolometers along a column, giving it a distinct
signature from other common signals that always appear with the same
sign.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{skynoise}
\caption{Illustration of the dominance of atmospheric variations on
  long-timescales using 35\,min. of data taken in highly-variable
  weather conditions (line-of-sight 450\,\micron\ opacity ranging from
  3.6 to 7.7) from the fully-commissioned instrument, observation 18
  on 2011-03-24. Red and blue lines show the 850 and 450\,\micron\
  common-mode signals, corresponding to the left- and right- vertical
  axes, respectively. The black line shows the line-of-sight opacity
  from the WVM as a proxy for atmospheric emission (arbitrary
  scale). Bolometer baseline drifts on timescales $\gsim$100\,s are
  clearly correlated with the atmosphere in this case.}
\label{fig:skynoise}
\end{figure}

The data shown in Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec} were
taken during fairly stable weather conditions (although with different
450\,\micron\ line-of-sight opacities $\sim$0.9 and 6.5 in the S2SRO
and post-upgrade data sets, respectively), and are dominated at
low frequencies ($\lsim$2\,Hz) by scan-synchronous noise, rather than
underlying variations in the atmosphere, or ``sky noise''. However,
only $\sim$100\,s of data were analysed. In Fig.~\ref{fig:skynoise},
an order-of-magnitude longer (fully-commissioned) data set is plotted
from observation 18 on 2011-03-24, during highly variable weather
conditions (450\,\micron\ line-of-sight opacities ranging from 3.6 to
7.7, scaled from the WVM data which are shown as a black line). Unlike
the earlier plots, it is clear that the bolometer baseline drifts on
timescales $\gsim$100\,s are tightly correlated with the atmospheric
opacity, which serves as a proxy for atmospheric emission. The weaker
correlation between the WVM and the 450\,\micron\ bolometers is
probably due to the atmosphere being more transmissive at 850\micron,
therefore the emission detected at 450\,\micron\ comes from water
vapour that is closer to the telescope.

While the relative importance of sky noise at low frequencies is
highly time- and scan pattern-dependent, in general it does not have a
major impact on map-making. Upon common-mode removal (which is an
integral part of our map-making strategy) uncorrelated noise remains
(e.g., solid lines in Fig.~\ref{fig:pspec}). As we will see in
Section~\ref{sec:pca} and \ref{sec:experimental}, this remaining noise
does not have a smoothly-varying correlation pattern across the focal
plane as one might expect if it were due to resolved ``clouds'' of
atmospheric emission. A high-pass filter with an edge frequency in the
range $\sim$0.1--1.0\,Hz is usually employed to remove this residual
noise (Section~\ref{sec:components}), and will also remove any sky
noise that is not already subtracted by the common-mode.

@@

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of
magnetic field pickup and brightness variations in the atmosphere)?
First, the smoothly-varying gradients in the eigenvalue maps
(primarily at 850\,\micron) suggests that this noise has a different
response across the focal plane; in this case the PCA has identified
two orthogonal shapes that, when mixed in different quantities, can
reproduce most of the signal in each bolometer. It is also likely that
the atmospheric contribution is stronger at 450\,\micron\ (i.e.,
predominantly Component~3).
